Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-13 paper/paper.tex
proof writing done, now for space saving

{\color{Myblue}  As noted  earlier, $x.\alpha  \in \Lv$  specifies the
  liveness of the reference $\heap\llbracket\rho(x), \alpha\rrbracket$
  only   if   $\alpha$  corresponds   to   a   closure-free  path   in
  \heap\ starting  from $\rho(x)$.  If  this path is intercepted  by a
  closure,  say  $(\CAR~{y_{\pi}})$, then  the  liveness  of the  path
  starting from $\mathit{y_{\pi}}$ is given by $\Lv_\mathit{y_{\pi}}$.
  As we shall see in  Section~\ref{sec:GC-scheme}, the liveness of the
  closure variable  $y_{\pi}$ is recorded  along with the  closure for
  $s$ so that the GC can refer to it during garbage collection.  }
%} %comment

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}. {\color{Myblue} It is easy to see that the analysis
correctly identifies the liveness of stack variables. A stack variable
is live
between its introduction through a \LET\  and its last use to
create a closure variable. This is correctly captured by the
\LET\ rule in Figure~\ref{fig:live-judge}.}

Proving soundness  for closure  variables is more complex. Here  are the
ideas behind the proof. 
\begin{enumerate}
\item      We     augment      the      standard     semantics      in
  Figure~\ref{fig:lang-semantics} to  model a GC before  the execution
  of  each \LET\  expression.  {\color{Myblue}Note  that, unlike  eager
    languages,   memory  is   allocated  only   during  execution   of
    \LET\ expressions}. During GC, we track each reference in the root
  set and heap that is declared  dead by our analysis.  Any attempt to
  dereference  such  references  results   in  the  transition  system
  entering a special state denoted \bang.  We call the semantics after
  augmentation, \emph{minefield semantics}.

@@

\hline 
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} 
\\
\hline
\makecell[t]{ $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \rho(x),
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, 
\cred{\renewcommand{\arraystretch}{1}\left\{
\begin{array}{lr}
  {\emptyset}&\mbox{if } \sigma = \emptyset \\       
  {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
\end{array}\right.
}$         
&
{\sc car-clo}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
{\sc let} \\ 
\hline
\end{tabular}
\end{center}
\caption{Minefield  semantics for  {\sc  car-bang}  and {\sc  car-clo}
  \warning{Add a note mentioning that we handle the issue of requiring
    a      specific     demand      by      introducing     a      new
    symbol}\label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to non-live references being replaced by $\bot$.  This simulates the
  act of  garbage collecting  the cells
  pointed to by these references  during an actual garbage collection.
  To do so, $GC$ needs to consider the following environments:  (1) the  environment in  the
  current state, (2) the  environment in each of the stacked  continuations and (3)
  the environment  in each of  the closures in  the heap.

 \begin{enumerate} 
{\color{Myblue}

 \item \label{env}  For each of  these environments, $GC$  calculates a
   liveness  environment $\Lv$  using  the  corresponding  $e$  (or  $s$)  and
   $\sigma$.
 \item 
   For each
    location $\ell$,  $GC$  sets $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ above, for each $x \in domain(\rho)$, and each forward
    access
    path $\alpha$, it is \emph{not} the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
}

@@

%%     For each
%%     location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
%%     environment $\rho$ in either  the  current state or in a
%%     continuation frame, for each $x \in domain(\rho)$ and each forward
%%     access
%%     path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$. Here 
%% %% $x.\alpha$  is  live  and
%% %%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%% %%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%% %%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
%% \Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
%% step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  GC
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

Consider a trace  of a minefield execution of a  program $p$, possibly
ending in  a \bang\ state. We  can construct a call-tree  based on the
trace in  which each node represents  a function that was  called (but
did not necessarily  return because of a \bang).  Assume  that each of
the nodes of  the tree is also annotated with  the program point where
the corresponding call  was invoked.  This tree can be  used to inline
function calls in a hierarchical fashion.  The details of the inlining
can be found in~\cite{asati14lgc}.
For  a
call-less program,  the initial state of  the minefield
semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
e_{\mainpgm}$.

@@

Note that our proofs will be for a single round of minefield execution
i.e.  the evaluation of $(\mainpgm)$ (or $e_{\mainpgm}$, for call-less
programs)  to   its  WHNF  driven  by   the  printing  mechanism~(Section
\ref{sec:semantics}).  With  minor variations,  the proof will  also be
applicable for subsequent evaluations initiated by $\print$.
%% We show  next that the  minefield execution of a  call-less program
%% cannot go \bang.

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
step. Consider a state $(\rho, S,  \heap, e, \sigma)$ in the minefield
execution of a  program.  We show by induction on  the number of steps
$n$ leading  to this  state that  the next  transition cannot  enter a
\bang\  state.    When  $n$   is  0,   the  state   is  $([\;]_\rho,\,
(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
        [\;]_{\heap},\, (\mainpgm))$.  Since the  call to $GC$ in this
        state  does   nothing,  we  ju1st   have  to  show   that  the
        $\rightsquigarrow$  transition cannot  enter  a \bang\  state.
        Since our  programs are in  ANF, $e_{\mainpgm}$ can only  be a
        $\LET$  expression.   A  {\sc   let}  step  does  not  involve
        dereferencing, and thus cannot result in a \bang.

For the inductive step, we shall show that none of the minefield steps
that involves dereferencing  results in a \bang.  These  are the steps
which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
inserted by the immediately preceding  garbage collection or the GC of
an earlier minefield  step.  However the demand $\sigma'$  on basis of
which an earlier  collection would have inserted a  $\bot$, would have
included the  current demand $\sigma$.   Thus, if the  earlier garbage
collection had inserted a $\bot$, so would the current GC.  Thus it is
enough to  consider the $\rightsquigarrow$  along with the  current GC
and show that it does not lead to a \bang.

We consider the rules {\sc car-clo} and {\sc car-bang} only.  
The rest  of the rules involve similar reasoning. For the {\sc
  car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
\sigma$, we  know that $\sigma$ is  non-null. Therefore
the  liveness  of  $x$  includes  $\epsilon$,  and  the
dereferencing $\heap(\rho(x))$ will go without \bang.

For the  {\sc car-bang}  rule, observe that  there are
two dereferences.  First $x$  is dereferenced to  get a
cons  cell  and then  the  head  of  the cons  cell  is
dereferenced  to  obtain  a  closure.   If  the  demand
$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
of    $x$   will    include    both   $\epsilon$    and
$\acar\epsilon$,  and  a  GC with  this  liveness  will
neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
first   component  of   the   cons   cell.  Thus   both
dereferences    can take place without 
$\rightsquigarrow$ entering the   \bang\ state.
\end{proof}

@@

%\subsection{Modifying liveness rules}
\paragraph{Symbolic representation of operations}

The   $\mathit{ref}$   rule   for   \CONS,   shown   in
Figure~\ref{fig:live-judge}, requires us  to remove the
leading  \acar\ and  \acdr\  from the  access paths  in
$\sigma$.  Similarly, the rules  for \CAR, \CDR, \PRIM,
\NULLQ, and \SIF\ require  us to return $\emptyset$, if
$\sigma$      itself       is      $\emptyset$      and
$\lbrace\epsilon\rbrace$  otherwise.  To  realize these
rules  $\sigma$   needs  to  be  known.   This  creates
difficulties  since  we  want to  solve  the  equations
arising from liveness symbolically.

@@

%\subsection{Generating equations for $\Lfonly_{\mathit f}$}
\paragraph{Computing function summaries $\Lfonly_{\mathit f}$}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2-a}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{x_i}{}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$.   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
$$
 \Lf{\length}{1}{\sigma} = \Lanv{\xl}{} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
$$

@@

%\subsection{Generating liveness equations \Lv\  for function bodies}
\paragraph{Handling user-defined functions}
\label{sec:bodylivenessbodies}

To  avoid analyzing  the body  of a  function for  each
call, we calculated the  liveness for the arguments and
the variables in a function  with respect to a symbolic
demand  $\sigma$.   To  get   the  actual  liveness  we
calculate an  over-approximation of the  actual demands
made by  all the  calls and  calculate the  liveness at
each  GC  point  inside  the  function  based  on  this
approximation.  The 0-CFA-style {\em summary demand} is
calculated by  taking a union  of the demands  at every
call site of a function.

@@

%\subsection{Solving liveness equations---the grammar interpretation}\label{sec:grammar-formulation}      
\subsection{From liveness sets to context-free grammars}\label{sec:grammar-formulation}

The  equations  above can  now  be  re-interpreted as  a  context-free
grammar  (CFG) on  the  alphabet $\lbrace\acar,  \acdr, \bcar,  \bcdr,
\clazy\rbrace$.  Let \var{$X$} denote  the non-terminal for a variable
$X$ occurring on the LHS of the equations generated from the analysis.
We can  think of  the resulting productions  as being  associated with
several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
as a start symbol.  As  an example, the grammar for \var{\Lanv{\xl}{}}
and \var{\Lanv{\pa}{}} comprise the following productions:
\begin{eqnarray*}
  \var{\Lanv{\xl}{}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{x}{}},   denoted
$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
solution  of  \Lanv{x}{}.    
However, recall that the decision problem that we are interested in
during GC is: 
%%\begin{quote} 
{\em
Let $x.\alpha$ be a forward access path (consisting only
of edges  \acar\ and \acdr\  but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{x}{}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
}
%%\end{quote}
We model  this problem as one  of deciding the
membership of  a CFG augmented  with a
fixed set of unrestricted productions.

>>>>

\added{  As noted  earlier, $x.\alpha  \in \Lv$  specifies the
  liveness of the reference $\heap\llbracket\rho(x), \alpha\rrbracket$
  only   if   $\alpha$  corresponds   to   a   closure-free  path   in
  \heap\ starting  from $\rho(x)$.  If  this path is intercepted  by a
  closure,  say  $(\CAR~{y_{\pi}})$, then  the  liveness  of the  path
  starting from $\mathit{y_{\pi}}$ is given by $\Lv_\mathit{y_{\pi}}$.
  As we shall see in  Section~\ref{sec:GC-scheme}, the liveness of the
  closure variable  $y_{\pi}$ is recorded  along with the  closure for
  $s$ so that the GC can refer to it during garbage collection.  }
%} %comment

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}. \added{It is easy to see that the analysis
correctly identifies the liveness of stack variables. A stack variable
is live
between its introduction through a \LET\  and its last use to
create a closure variable. This is correctly captured by the
\LET\ rule in Figure~\ref{fig:live-judge}.}
Proving soundness  for closure  variables is more complex. Here  are the
ideas behind the proof. 
\begin{enumerate}
\item      We     augment      the      standard     semantics      in
  Figure~\ref{fig:lang-semantics} to  model a GC before  the execution
  of  each \LET\  expression.  \added{Note  that, unlike  eager
    languages,   memory  is   allocated  only   during  execution   of
    \LET\ expressions}. During GC, we track each reference in the root
  set and heap that is declared  dead by our analysis.  Any attempt to
  dereference  such  references  results   in  the  transition  system
  entering a special state denoted \bang.  We call the semantics after
  augmentation, \emph{minefield semantics}.

@@

\hline 
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} 
\\
\hline
\makecell[t]{ $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \rho(x),
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, 
\cred{\renewcommand{\arraystretch}{1}\left\{
\begin{array}{lr}
  {\emptyset}&\mbox{if } \sigma = \emptyset \\       
  {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
\end{array}\right.
}$         
&
{\sc car-clo}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
{\sc let} \\ 
\hline
\end{tabular}
\end{center}
\caption{Minefield  semantics for  \CAR\   and \LET.
  \warning{Add a note mentioning that we handle the issue of requiring
    a      specific     demand      by      introducing     a      new
    symbol}\label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$. The changes in
  $\rho, S$ and  $\heap$ are due
  to non-live references being replaced by $\bot$.  This simulates the
  act of  garbage collecting  the cells
  pointed to by these references  during an actual garbage collection.
  To do so, $GC(\ldots)$ needs to consider the following environments:  (1) the  environment in  the
  current state, (2) the  environment in each of the stacked  continuations and (3)
  the environment  in each of  the closures in  the heap.

 \begin{enumerate} 
\added{

 \item \label{env}  For each of  these environments, $GC(\ldots)$  calculates a
   liveness  environment $\Lv$  using  the  corresponding  $e$  (or  $s$)  and
   $\sigma$.
 \item 
   For each
    location $\ell$,  $GC(\ldots)$  sets $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ above, for each $x \in domain(\rho)$, and each forward
    access
    path $\alpha$, it is \emph{not} the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
}

@@

%%     For each
%%     location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
%%     environment $\rho$ in either  the  current state or in a
%%     continuation frame, for each $x \in domain(\rho)$ and each forward
%%     access
%%     path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$. Here 
%% %% $x.\alpha$  is  live  and
%% %%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%% %%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%% %%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
%% \Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
%% step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Figure ~\ref{fig:minefield-semantics-for-some} shows some of the
minefield rules. As mentioned earlier, the transition for a \LET\ 
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
is preceded  by $GC(\ldots)$.  The details of the transition for the {\sc
  car-clo}  rule   is  also shown.  If    an
earlier  call to  $GC(\ldots)$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

Consider a trace  of a minefield execution of a  program $p$, possibly
ending in  a \bang\ state. We  can construct a call-tree  based on the
trace in  which each node represents  a function that was  called (but
did not necessarily  return because of a \bang).  Assume  that each of
the nodes of  the tree is also annotated with  the program point where
the corresponding call  was invoked.  This tree can be  used to inline
function calls in a hierarchical fashion.  The details of the inlining
can be found in~\cite{asati14lgc}.
For  a
call-less program,  the initial state of  the minefield
semantics  is  assumed  to  be  $([\;]_\rho,(\rho_\mathit{init},
\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
e_{\mainpgm})$.

@@

Note that our proofs will be for a single round of minefield execution
i.e.  the evaluation of  $(\mainpgm)$  to   its  WHNF  driven  by   the  printing  mechanism~(Section
\ref{sec:semantics}).  With  minor variations,  the proof will  also be
applicable for subsequent evaluations initiated by $\print$.
%% We show  next that the  minefield execution of a  call-less program
%% cannot go \bang.

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Consider a  state  $(\rho,  S,  \heap,  e, \sigma)$  in  the
minefield execution of a program.  We  show by induction on the number $n$
of $\rightsquigarrow$ steps  leading to  this state  that the next  transition cannot
enter a  \bang\ state.   When $n$  is 0,  the state  is $([\;]_\rho,\,
(\rho_\mathit{init},   \,  \ell_{\ans}   ,\,  (\print~\ans)):[\;]_{S},
[\;]_{\heap},\, e_{\mainpgm})$.   Since the call  to $GC(\ldots)$ in  this state
does  nothing,  we  just  have to  show  that  the  $\rightsquigarrow$
transition cannot  enter a  \bang\ state.  Since  our programs  are in
ANF, $e_{\mainpgm}$ can only be a $\LET$ expression.  A {\sc let} step
does not involve dereferencing, and thus cannot result in a \bang.

For the inductive step, we shall show that none of the minefield steps
that involves dereferencing  results in a \bang.  These  are the steps
which   have   a   $\heap(\rho(...))$   in   the   premise.    Now   a
$\rightsquigarrow$ step can go \bang\ because it dereferences a $\bot$
inserted by an earlier $GC(\ldots)$.   However the demand $\sigma'$  on basis of
which the $GC(\ldots)$ would have inserted a  $\bot$ would have
included the  current demand $\sigma$.     Thus it is
enough to  show that the $\rightsquigarrow$ step would not lead to a
\bang, even if $GC(\ldots)$ had been done with the current demand
$\sigma$.

We consider the rules for \CAR\  only.  
The rest  of the rules involve similar reasoning. For the {\sc
  car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
\sigma$, we  know that $\sigma$ is  non-null. Therefore
the  liveness  of  $x$  includes  $\epsilon$,  and  the
dereferencing $\heap(\rho(x))$ will go without \bang.

For the  {\sc car-1-clo}  rule, observe that  there are
two dereferences.  First $x$  is dereferenced to  get a
cons  cell  and then  the  head  of  the cons  cell  is
dereferenced  to  obtain  a  closure.   If  the  demand
$\sigma$ on  $(\CAR~x)$ is non-null, then  the liveness
of    $x$   will    include    both   $\epsilon$    and
$\acar\epsilon$,  and  a  GC with  this  liveness  will
neither bind $x$ to a  $\bot$, nor insert $\bot$ at the
first   component  of   the   cons   cell.  Thus   both
dereferences    can take place without 
$\rightsquigarrow$ entering the   \bang\ state.
\end{proof}

@@

%\subsection{Modifying liveness rules}
\paragraph{Symbolic representation of operations:}
The   $\mathit{ref}$   rule   for   \CONS,   shown   in
Figure~\ref{fig:live-judge}, requires us  to remove the
leading  \acar\ and  \acdr\  from the  access paths  in
$\sigma$.  Similarly, the rules  for \CAR, \CDR, \PRIM,
\NULLQ, and \SIF\ require  us to return $\emptyset$, if
$\sigma$      itself       is      $\emptyset$      and
$\lbrace\epsilon\rbrace$  otherwise.  To  realize these
rules  $\sigma$   needs  to  be  known.   This  creates
difficulties  since  we  want to  solve  the  equations
arising from liveness symbolically.

@@

%\subsection{Generating equations for $\Lfonly_{\mathit f}$}
\paragraph{Computing function summaries $\Lfonly_{\mathit f}$:}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2-a}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{x_i}{}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$.   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
$$
 \Lf{\length}{1}{\sigma} = \Lanv{\xl}{} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
$$

@@

%\subsection{Generating liveness equations \Lv\  for function bodies}
\paragraph{Handling user-defined functions:}
\label{sec:bodylivenessbodies}
To  avoid analyzing  the body  of a  function for  each
call, we calculated the  liveness for the arguments and
the variables in a function  with respect to a symbolic
demand  $\sigma$.   To  get   the  actual  liveness  we
calculate an  over-approximation of the  actual demands
made by  all the  calls and  calculate the  liveness at
each  GC  point  inside  the  function  based  on  this
approximation.  The 0-CFA-style {\em summary demand} is
calculated by  taking a union  of the demands  at every
call site of a function.

@@

%\subsection{Solving liveness equations---the grammar interpretation}\label{sec:grammar-formulation}      
\paragraph{From liveness sets to context-free grammars:}\label{sec:grammar-formulation}      
The  equations  above can  now  be  re-interpreted as  a  context-free
grammar  (CFG) on  the  alphabet $\lbrace\acar,  \acdr, \bcar,  \bcdr,
\clazy\rbrace$.  Let \var{$X$} denote  the non-terminal for a variable
$X$ occurring on the LHS of the equations generated from the analysis.
We can  think of  the resulting productions  as being  associated with
several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
as a start symbol.  As  an example, the grammar for \var{\Lanv{\xl}{}}
and \var{\Lanv{\pa}{}} comprise the following productions:
\begin{eqnarray*}
  \var{\Lanv{\xl}{}}  &\rightarrow& 
  \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
  \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\epsilon  \mid \clazy\langle{\sigma_{\length}}\rangle \\
\var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{x}{}},   denoted
$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
solution  of  \Lanv{x}{}.    
However, recall that the decision problem that we are interested in
during GC is: 
%%\begin{quote} 
{\em
Let $x.\alpha$ be a forward access path (consisting only
of edges  \acar\ and \acdr\  but not \bcar,  \bcdr\ or
\clazy).       Let      $\mathscr{L}(\var{\Lanv{x}{}})
\stackrel{*}{\hookrightarrow}  \sigma$, where  $\sigma$
consists  of  forward  access  paths  only.  Then  does
$\alpha \in \sigma$?
}
%%\end{quote}
We model  this problem as one  of deciding the
membership of  a CFG augmented  with a
fixed set of unrestricted productions.
